A quantum parallel Markov chain Monte Carlo

### 2023-01-25 manuscript/main.tex
fixed small error
more edits

@@

\begin{figure}[!t]
	\centering
	\scalebox{0.9}{
\begin{minipage}{1\textwidth}
\begin{algorithm}[H]
	\caption{Parallel MCMC \citep{tjelmeland2004using}}\label{alg:pMCMC}
	\KwData{Initial Markov chain state $\ttheta^{(0)}$; total length of Markov chain $S$; total number of proposals per iteration $P$; routine for evaluating target density $\pi(\cdot)$; routines for drawing random samples from the proposal distribution $Q(\ttheta^{(s)},\cdot)$ and from a $P+1$ discrete distribution $Discrete(\ppi)$ given some probability vector $\ppi=(\pi_0,\dots,\pi_P)$.}
	\KwResult{A Markov chain $\ttheta^{(1)}, \dots, \ttheta^{(S)}$.}
	\For{$s \in\{1,\dots,S\}$}{
		%$\sigma_\pi \gets 0$\;
		$\ttheta_0 \gets \ttheta^{(s-1)}$\;
		$\Ttheta_{-0} \gets Q(\ttheta_0,\cdot)$\;
		\For{$p \in \{0,\dots,P\}$}{
			$\pi_{p} \gets \pi(\ttheta_{p})\, q(\ttheta_{p},\Ttheta_{-p})$\; 
		}
		$\hat{p} \gets Discrete(\ppi/\ppi^T\mathbb{1})$\;
		$\ttheta^{(s)} \gets \ttheta_{\hat{p}}$\;
	}
	\Return{$\ttheta^{(1)}, \dots, \ttheta^{(S)}$}\ .
	\vspace{0.5em}

@@

While parallel MCMC algorithms are increasingly well-suited for developing many-core computational architectures, there are trade-offs that need to be taken into account when choosing how to allocate computational resources.  On one end of the spectrum, \citet{gelman1992inference} demonstrate the usefulness of generating, combining and comparing multiple independent Markov chains that target the same distribution, and one may perform this embarrassingly parallel task by assigning the operations for each individual chain to a separate central processing unit (CPU) core or GPU work-group.  In this multi-chain context, simultaneously running multiple parallel MCMC chains could limit resources available for the within-chain parallelization described in the previous paragraph.  On the other end of the spectrum, one may find it useful to allocate resources to overcome computational bottlenecks within a single Markov chain that uses a traditional accept-reject step. In big data contexts, \citet{massive,holbrook2021scalable,holbrook2022bayesian,holbrook2022viral} use multi- and many-core processing to accelerate log-likelihood and log-likelihood gradient calculations within single Metropolis-Hastings and Hamiltonian Monte Carlo \citep{neal2011mcmc} generated chains.  This big data strategy might again limit resources available for parallelization across proposals and target evaluations described above.

@@

\begin{figure}[!t]
	\centering
	\scalebox{0.9}{
		\begin{minipage}{1\textwidth}
			\begin{algorithm}[H]
	\caption{The Gumbel-max trick}\label{alg:gm}
	\KwData{A vector of unnormalized log-probabilities $\llambda^*=\log \ppi +\log c$, for $\ppi$ a discrete probability vector with $P+1$ elements.}
	\KwResult{A single sample $\hat{p} \sim$ \emph{Discrete}$(\ppi)$ satisfying $\hat{p}\in\{0,1,\dots,P\}$.}
	\For{$p \in\{0,1,\dots,P\}$}{
		$z_{p} \gets Gumbel(0,1)$\;
		$\alpha^*_{p} \gets \lambda_{p}^* + z_{p}$\;
	}
	$\hat{p} \gets \argmax_{p=0,\dots,P}\alpha^*_{p}$\;
	\Return{p}\ .

@@

\begin{figure}[!t]
	\centering
	\scalebox{0.9}{
		\begin{minipage}{1\textwidth}
			\begin{algorithm}[H]
	\caption{Quantum parallel MCMC for a continuously-valued target}\label{alg:qpMCMC}
	\KwData{Initial Markov chain state $\ttheta^{(0)}$; total length of Markov chain $S$; total number of proposals per iteration $P$; routine for evaluating target density $\pi(\cdot)$; routines for drawing random samples from a $D$-dimensional multivariate Gaussian $Normal_D(\boldsymbol{\mu},\SSigma)$ and the standard Gumbel distribution $Gumbel(0,1)$.}
	\KwResult{A Markov chain $\ttheta^{(1)}, \dots, \ttheta^{(S)}$.}
	\For{$s \in\{1,\dots,S\}$}{
		$\ttheta_0 \gets \ttheta^{(s-1)}$\;
		$z_{0} \gets Gumbel(0,1)$\;
		$\bar{\ttheta} \gets Normal_D(\ttheta_0,\SSigma)$; \hspace{14em} \Comment{ \hspace{0.3em}     Proposal \eqref{eq:p1}}      
		\For{$p \in \{1,\dots,P\}$}{
			$\ttheta_p \gets Normal_D(\bar{\ttheta},\SSigma)$\;
			$z_{p} \gets Gumbel(0,1)$\;
			$\ket{\theta_p} \gets \ttheta_p$; \hspace{5em} \Comment{      Load proposal onto quantum computer.}      
		}
		$\hat{p} \gets \argmin_{p=0,\dots,P}\Big(f(p):= -\big(z_{p}+ \log\pi(\ttheta_{p}) \big)\Big)$\;  \hspace{10em}\Comment{\hspace{0.5em} $\order{\sqrt{P}}$ Alg \ref{alg:min} with warm-start at $p=0$.}
		$\ttheta^{(s)} \gets \ttheta_{\hat{p}}$\;
	}
	\Return{$\ttheta^{(1)}, \dots, \ttheta^{(S)}$}\ .

@@

Algorithm \ref{alg:qpMCMC} presents the details of QPMCMC for a continuously-valued target.  The algorithm uses conventional (perhaps parallel) computing almost entirely, excluding the one line that calls the quantum minimization algorithm presented in Algorithm \ref{alg:min}. 
This sub-routine seeks to find the minimizer of the function
\begin{align}\label{eq:objective}
f(p)= -\big(z_{p}+ \log\pi(\ttheta_{p}) \big) \quad \mbox{over} \quad p=0,1,\dots, P,
\end{align} 
which combines the numerators of \eqref{eq:probs}, the Gumbel-max trick and a trivial negation.  As discussed in Section \ref{sec:qmin}, the sub-routine requires only $\mathcal{O}(\sqrt{P})$ oracle evaluations and, therefore, only $\mathcal{O}(\sqrt{P})$ evaluations of the target density $\pi(\cdot)$ using a quantum circuit.  In theory, a quantum computer can perform the same computations as a classical computer, but the efficiency of the target evaluations would depend on a number of factors related to the structure of the density itself, the availability of fast quantum random access memory \citep{giovannetti2008quantum} and the ability of large numbers of quantum gates to act in concert with negligible noise \citep{kielpinski2002architecture,erhard2019characterizing}.

@@

To ascertain whether QPMCMC mixes differently compared to conventional parallel MCMC, we run both algorithms for a range of proposal counts to sample from a 10-dimensional standard normal distribution. For each algorithm and proposal setting, we run 100 independent chains for 10,000 iterations and obtain effective sample sizes ESS$_d$ for $d \in \{1,\dots,10\}$.   We then calculate the relative differences between the means and minima of one chain generated using parallel MCMC and one chain generated using QPMCMC; for example:
\begin{align*}
	\mbox{Relative difference between means} \: \overline{\mbox{ESS}}_{(\cdot)} = \frac{\left|\overline{\mbox{ESS}}_{pMCMC} - \overline{\mbox{ESS}}_{QPMCMC}\right| }{\overline{\mbox{ESS}}_{pMCMC} }
\end{align*}
Figure \ref{fig:mixing} shows results.  In general, the majority relative differences are small.  For both statistics, mean relative differences are less than 0.5, regardless of proposal count.  Again for both statistics, more than 75\% of the independent runs result in relative differences below 0.1.  We note that relative differences between means (blue) appear to decrease with the number of proposals, but the same does not hold for relative differences between minima (red).

>>>>

\def\HiLi{\leavevmode\rlap{\hbox to \hsize{\color{trevorblue!15}\leaders\hrule height 0.9\baselineskip depth .5ex\hfill}}}

@@

\begin{figure}[!t]
	\centering
	\scalebox{0.9}{
\begin{minipage}{1\textwidth}
\begin{algorithm}[H]
	\caption{Parallel (multiproposal) MCMC \citep{tjelmeland2004using}}\label{alg:pMCMC}
	\KwData{Initial Markov chain state $\ttheta^{(0)}$; total length of Markov chain $S$; total number of proposals per iteration $P$; routine for evaluating target density $\pi(\cdot)$; routines for drawing random samples from the proposal distribution $Q(\ttheta^{(s)},\cdot)$ and from a $P+1$ discrete distribution $Discrete(\ppi)$ given some probability vector $\ppi=(\pi_0,\dots,\pi_P)$.}
	\KwResult{A Markov chain $\ttheta^{(1)}, \dots, \ttheta^{(S)}$.}
	\For{$s \in\{1,\dots,S\}$}{
		%$\sigma_\pi \gets 0$\;
		$\ttheta_0 \gets \ttheta^{(s-1)}$\;
		$\Ttheta_{-0} \gets Q(\ttheta_0,\cdot)$\;
		\For{$p \in \{0,\dots,P\}$}{
			$\pi_{p} \gets \pi(\ttheta_{p})\, q(\ttheta_{p},\Ttheta_{-p})$\; 
		}
		$\hat{p} \gets Discrete(\ppi/\ppi^T\mathbb{1})$\;
		$\ttheta^{(s)} \gets \ttheta_{\hat{p}}$\;
	}
	\Return{$\ttheta^{(1)}, \dots, \ttheta^{(S)}$}\ .
	\vspace{0.5em}

@@

While parallel MCMC algorithms are increasingly well-suited for developing many-core computational architectures, there are trade-offs that need to be taken into account when choosing how to allocate computational resources.  On one end of the spectrum, \citet{gelman1992inference,gelman1992single} demonstrate the usefulness of generating, combining and comparing multiple independent Markov chains that target the same distribution, and one may perform this embarrassingly parallel task by assigning the operations for each individual chain to a separate central processing unit (CPU) core or GPU work-group.  In this multi-chain context, simultaneously running multiple parallel MCMC chains could limit resources available for the within-chain parallelization described in the previous paragraph.  On the other end of the spectrum, one may find it useful to allocate resources to overcome computational bottlenecks within a single Markov chain that uses a traditional accept-reject step. In big data contexts, \citet{massive,holbrook2021scalable,holbrook2022bayesian,holbrook2022viral} use multi- and many-core processing to accelerate log-likelihood and log-likelihood gradient calculations within single Metropolis-Hastings and Hamiltonian Monte Carlo \citep{neal2011mcmc} generated chains.  This big data strategy might again limit resources available for parallelization across proposals and target evaluations described above.

@@

\begin{figure}[!t]
	\centering
	\scalebox{0.9}{
		\begin{minipage}{1\textwidth}
			\begin{algorithm}[H]
	\caption{The Gumbel-max trick}\label{alg:gm}
	\KwData{A vector of unnormalized log-probabilities $\llambda^*=\log \ppi +\log c$, for $\ppi$ a discrete probability vector with $P+1$ elements.}
	\KwResult{A single sample $\hat{p} \sim$ \emph{Discrete}$(\ppi)$ satisfying $\hat{p}\in\{0,1,\dots,P\}$.}
	\For{$p \in\{0,1,\dots,P\}$}{
		$z_{p} \gets Gumbel(0,1)$\;
		$\alpha^*_{p} \gets \lambda_{p}^* + z_{p}$\;
	}
	$\hat{p} \gets \argmax_{p=0,\dots,P}\alpha^*_{p}$\;
	\Return{$\hat{p}\,$}.

@@

\begin{figure}[!t]
	\centering
	\scalebox{0.9}{
		\begin{minipage}{1\textwidth}
			\begin{algorithm}[H]
	\caption{Quantum parallel MCMC for a continuously-valued target}\label{alg:qpMCMC}
	\KwData{Initial Markov chain state $\ttheta^{(0)}$; total length of Markov chain $S$; total number of proposals per iteration $P$; routine for evaluating target density $\pi(\cdot)$; routines for drawing random samples from a $D$-dimensional multivariate Gaussian $Normal_D(\boldsymbol{\mu},\SSigma)$ and the standard Gumbel distribution $Gumbel(0,1)$.}
	\KwResult{A Markov chain $\ttheta^{(1)}, \dots, \ttheta^{(S)}$.}
	\For{$s \in\{1,\dots,S\}$}{
		$\ttheta_0 \gets \ttheta^{(s-1)}$\;
		$z_{0} \gets Gumbel(0,1)$\;
		$\bar{\ttheta} \gets Normal_D(\ttheta_0,\SSigma)$; \hspace{14em} \Comment{ \hspace{0.3em}     Proposal \eqref{eq:p1}}      
		\For{$p \in \{1,\dots,P\}$}{
			$\ttheta_p \gets Normal_D(\bar{\ttheta},\SSigma)$\;
			$z_{p} \gets Gumbel(0,1)$\;
		\HiLi	$\ket{\theta_p} \gets \ttheta_p$; \hspace{5em} \Comment{      Load proposal onto quantum computer.}      
		}
	\HiLi	$\hat{p} \gets \argmin_{p=0,\dots,P}\Big(f(p):= -\big(z_{p}+ \log\pi(\ttheta_{p}) \big)\Big)$\;  \HiLi \hspace{10em}\Comment{\hspace{0.5em} $\order{\sqrt{P}}$ Alg \ref{alg:min} with warm-start at $p=0$.}
		$\ttheta^{(s)} \gets \ttheta_{\hat{p}}$\;
	}
	\Return{$\ttheta^{(1)}, \dots, \ttheta^{(S)}$}\ .

@@

Algorithm \ref{alg:qpMCMC} presents the details of QPMCMC for a continuously-valued target.  The algorithm uses conventional (perhaps parallel) computing almost entirely, excluding two lines that are highlighted.  The first of these loads proposals onto the quantum machine, and the second line that calls the quantum minimization algorithm presented in Algorithm \ref{alg:min}. 
This sub-routine seeks to find the minimizer of the function
\begin{align}\label{eq:objective}
f(p)= -\big(z_{p}+ \log\pi(\ttheta_{p}) \big) \quad \mbox{over} \quad p=0,1,\dots, P,
\end{align} 
which combines the numerators of \eqref{eq:probs}, the Gumbel-max trick and a trivial negation.  As discussed in Section \ref{sec:qmin}, the sub-routine requires only $\mathcal{O}(\sqrt{P})$ oracle evaluations and, therefore, only $\mathcal{O}(\sqrt{P})$ evaluations of the target density $\pi(\cdot)$ using a quantum circuit.  In theory, a quantum computer can perform the same computations as a classical computer, but the efficiency of the target evaluations would depend on a number of factors related to the structure of the density itself, the availability of fast quantum random access memory \citep{giovannetti2008quantum} and the ability of large numbers of quantum gates to act in concert with negligible noise \citep{kielpinski2002architecture,erhard2019characterizing}.

@@

To ascertain whether QPMCMC mixes differently compared to conventional parallel MCMC, we run both algorithms for a range of proposal counts to sample from a 10-dimensional standard normal distribution. For each algorithm and proposal setting, we run 100 independent chains for 10,000 iterations and obtain effective sample sizes ESS$_d$ for $d \in \{1,\dots,10\}$.   We then calculate the relative differences between the means and minima of one chain generated using parallel MCMC and one chain generated using QPMCMC; for example:
\begin{align*}
	\mbox{Relative difference between means} \: \overline{\mbox{ESS}}_{(\cdot)} = \frac{\left|\overline{\mbox{ESS}}_{pMCMC} - \overline{\mbox{ESS}}_{QPMCMC}\right| }{\overline{\mbox{ESS}}_{pMCMC} }
\end{align*}
Figure \ref{fig:mixing} shows results.  In general, the majority relative differences are small.  For both statistics, mean relative differences are less than 0.05, regardless of proposal count.  Again for both statistics, more than 75\% of the independent runs result in relative differences below 0.1.  We note that relative differences between means (blue) appear to decrease with the number of proposals, but the same does not hold for relative differences between minima (red).

### 2023-01-25 manuscript/revision_letter.tex
fixed small error
more edits

Finally, I've added Section \ref{sec:mixing}, which ompares mixing between conventional parallel MCMC and QPMCMC:

@@

	Figure \ref{fig:mixing} shows results.  In general, the majority relative differences are small.  For both statistics, mean relative differences are less than 0.5, regardless of proposal count.  Again for both statistics, more than 75\% of the independent runs result in relative differences below 0.1.  We note that relative differences between means (blue) appear to decrease with the number of proposals, but the same does not hold for relative differences between minima (red).

@@

\begin{itemize}
	\item (I am assuming the following is true.) It would be helpful to state clearly at some point that the resulting QPMCMC chain has the same transition kernel as the non-quantum version. That is, it remains exact.

	\item Equation (1): It is not clear to me how the ``a.s.'' applies to this definition of 1
	stationarity.

@@

	\item page 4, line 10: I believe the citation for \citet{gelman1992inference} is incorrect. Did the author mean \citet{gelman1992inference}?

@@

@@

@@

@@

	\item page 22: The authors tune to 50\% acceptance rate, but I believe the optimal in high- dimensions from the reference \citet{rosenthal2011optimal} is 23.4\%. Further, the tuning to 23.4\% acceptance rate is strictly for the Metropolis-Hastings acceptance function. In the multiple-proposal world, the acceptance function employed is completely different, implying different optimality results. See \citet{agrawal2021optimal}.

	\item Figure 5: ``Over 99.4\% of the 10,000 MCMC iterations successfully sampled from the discrete distribution with probabilities of Equation (2)''. It’s unclear exactly how this number comes about and what it means. Once the quantum minimization returns a $\hat{p}$, the authors have shown that it follows the right distribution. How can then (1) .6\% not be from the right distribution, and (2) how does one even quantify this.

	\item Figure 6: Considering the target is known, burn-in seems incredibly wasteful. Once can just start from stationarity and remove the need to throw away samples.

	\item Table 1: In producing this table, what effective sample size is being used for this two-dimensional problem?

	\item Could the authors comment on the potential applicability of the algorithm on discrete state spaces? This may be useful for Bayesian record linkage, and variable selection problems \citep{zanella2020informed}.

>>>>

Finally, I've added Section \ref{sec:mixing}, which compares mixing between conventional parallel MCMC and QPMCMC:

@@

	Figure \ref{fig:mixing} shows results.  In general, the majority relative differences are small.  For both statistics, mean relative differences are less than 0.05, regardless of proposal count.  Again for both statistics, more than 75\% of the independent runs result in relative differences below 0.1.  We note that relative differences between means (blue) appear to decrease with the number of proposals, but the same does not hold for relative differences between minima (red).

@@

\begin{itemize}
	\item (I am assuming the following is true.) It would be helpful to state clearly at some point that the resulting QPMCMC chain has the same transition kernel as the non-quantum version. That is, it remains exact.
\end{itemize}

\begin{reply}
	Sometimes Algorithm \ref{alg:min} does not find the minimum.  Because of this the QPMCMC chain does not have the same transition kernel as the non-quantum version.  Nonetheless, results like those in Figure \ref{fig:qq} show that bias remains small.  The second paragraph of Section \ref{sec:disc} reads:
	\begin{quote}
This work leads to three interesting questions.   First, what is the status of inference obtained by QPMCMC?  The QPMCMC selection step relies on quantum minimization, an algorithm that only achieves success with probability $1-\epsilon$.  While empirical studies suggest that this error induces little bias, it would be helpful to use this $\epsilon$ to bound the distance between the target distribution and the distribution generated by QPMCMC.  Such theoretical efforts would need to extend recent formalizations of the multiproposal based parallel MCMC paradigm \citep{glatt} to account for biased MCMC kernels.
	\end{quote}
I am currently working with collaborators to bound the Wasserstein distance between the target and the distribution actually sampled.  At the moment, I must leave such results for future work. 
\end{reply}

\begin{itemize}	
	\item Equation (1): It is not clear to me how the ``a.s.'' applies to this definition of 1
	stationarity.
\end{itemize}

\begin{reply}
	I thank the AE for catching this silly mistake, which  I have fixed.
\end{reply}

\begin{itemize}

@@

\begin{reply}
	This caption uses the term `parallel MCMC' in the same way that it is used in \citet{calderhead2014general}, \citet{schwedes2019parallel}, \citet{schwedes2021rao} and \citet{glatt}, i.e., to indicate the use of multiple proposals.  To help resolve the potential confusion, I have changed the caption to `Parallel (multiproposal) MCMC' and added the following text to Section \ref{sec:intro}:  
	\begin{quote}
		One need not use parallel computing to implement Algorithm \ref{alg:pMCMC}, but the real promise and power of parallel MCMC comes from its natural parallelizability \citep{calderhead2014general}. 
	\end{quote}
\end{reply}

\begin{itemize}

	\item page 4, line 10: I believe the citation for \citet{gelman1992inference} is incorrect. Did the author mean \citet{gelman1992single}?

\begin{reply}
	I really did mean \citet{gelman1992inference}, but I have now changed the text to cite both papers following the AE's helpful recommendation.  All I'm saying here is that there are different kinds of parallelization available, and that computational statisticians should engage the entire spectrum.  
\end{reply}

\begin{itemize}

@@

\begin{reply}
	I thank the AE for catching this, and I have changed the text accordingly.
\end{reply}

\begin{itemize}

@@

\begin{reply}
	I have moved the proof to Section \ref{sec:gm} of the Appendix.
\end{reply}

\begin{itemize}

@@

\begin{reply}
	I have highlighted the lines accordingly.  Also, I have edited the relevant text to read:
	\begin{quote}
		Algorithm \ref{alg:qpMCMC} presents the details of QPMCMC for a continuously-valued target.  The algorithm uses conventional (perhaps parallel) computing almost entirely, excluding two lines that are highlighted.  The first of these loads proposals onto the quantum machine, and the second line that calls the quantum minimization algorithm presented in Algorithm \ref{alg:min}. 
	\end{quote}
\end{reply}

\begin{itemize}

@@

\begin{itemize}

	\item page 22: The authors tune to 50\% acceptance rate, but I believe the optimal in high- dimensions from the reference \citet{rosenthal2011optimal} is 23.4\%. Further, the tuning to 23.4\% acceptance rate is strictly for the Metropolis-Hastings acceptance function. In the multiple-proposal world, the acceptance function employed is completely different, implying different optimality results. See \citet{agrawal2021optimal}.
\end{itemize}

\begin{itemize}

	\item Figure 5: ``Over 99.4\% of the 10,000 MCMC iterations successfully sampled from the discrete distribution with probabilities of Equation (2)''. It’s unclear exactly how this number comes about and what it means. Once the quantum minimization returns a $\hat{p}$, the authors have shown that it follows the right distribution. How can then (1) .6\% not be from the right distribution, and (2) how does one even quantify this.
\end{itemize}

\begin{itemize}	
	\item Figure 6: Considering the target is known, burn-in seems incredibly wasteful. Once can just start from stationarity and remove the need to throw away samples.
\end{itemize}

\begin{itemize}	
	\item Table 1: In producing this table, what effective sample size is being used for this two-dimensional problem?
\end{itemize}

\begin{itemize}	
	\item Could the authors comment on the potential applicability of the algorithm on discrete state spaces? This may be useful for Bayesian record linkage, and variable selection problems \citep{zanella2020informed}.
